The Effects of the NBA COVID Bubble on the NBA Playoffs: A Case Study for Home-Court Advantage

### 2022-02-13 Manuscript/manuscript.tex
Update manuscript.tex
Update manuscript.tex
Addition of Jones and Boudreaux sources and slight restructing of literature review

Some studies focus less on shooting and more on other metrics.
For example, \citet{Greer} focused on the influence of spectator booing on
home-court advantage in basketball. The three methods of performance used
in this study were scoring, violations, and turnovers. This study was conducted
using the men's basketball programs at two large universities. The study finds that
social support, like booing, is an important contributor to home-court advantage.
Greer explains, whether the influence is greater on visiting team performance or
referee calls is less clear. However, the data does seem to lean slightly in
favor of affecting visiting team performance. \citet{Harville} studied the
effect of home-court advantage using the 1991-1992 college basketball season.
Unlike the NBA, it is not uncommon to have a few games played at neutral sites
during the college basketball season. This allowed them to construct two samples,
one of home teams and one of neutral teams. They formulated their study in a
regression predicting the expected difference in score for home teams.
They set up their study to find if the home teams won games by more points when
they had home-court advantage vs.~when playing at neutral court. This study
concluded with evidence supporting home-court advantage.

@@

@@

It's worth noting that much of this paper discusses the effects from an
offensive point of view, focusing on a team's ability to score. However, one of 
the main factors that affects an offense's ability to score is the opposing
defense. It's possible that home-court advantage mainly functions as an extra
defender for home teams, which is how it negatively impacts away scoring. There
is not much home fans can do to help their team on offense, except being quiet
so players can easily communicate and focus. However, when on defense, fans can
act as a "sixth defender" supporting their team by cheering loudly and making it
harder on the offense. This would explain how away teams were able to improve
offensively in the bubble. Home teams did not have the regular help from the
noise of fans that hinders the away team's offensive efforts. This would also
explain why home team offense didn't show significant change in the bubble, the
in-game offensive environment was largely unaltered for home teams since the
bubble provided the same quiet environment they're used to at home while on
offense. I still believe home-court advantage is better categorized as a
negative effect on away team offense, since it's more likely
noise from fans lowers level of play of away offense rather than raises the level
of play of home defense. This will be further discussed below.

>>>>

Some studies focus less on shooting and more on other metrics.
For example, \citet{Greer} focused on the influence of spectator booing on
home-court advantage in basketball. The three methods of performance used
in this study were scoring, violations, and turnovers. This study was conducted
using the men's basketball programs at two large universities. The study finds that
social support, like booing, is an important contributor to home-court advantage.
Greer explains, whether the influence is greater on visiting team performance or
referee calls is less clear. However, the data does seem to lean slightly in
favor of affecting visiting team performance. Another study focused on scoring
was conducted by \cite{Jones}. \cite{Jones} analyzed scoring patterns across each of the 
4 quarters to analyze the effects of home-court advantage over the course of a game.
He found that typically 2/3 of the benefits of home-court advantage are received in 
the first quarter with the remaining 1/3 slowly accumulating over the 
final 3 quarters. This implies home-court advantage is most effective when teams 
use it to build a lead they then just have to maintain for the remainder of the game.
If they are losing after the first quarter, there is a sharp drop in win probability.

@@

There are a few examples of natural experiments in basketball. \citet{Harville} studied the
effect of home-court advantage using the 1991-1992 college basketball season.
Unlike the NBA, it is not uncommon to have a few games played at neutral sites
during the college basketball season. This allowed them to construct two samples,
one of home teams and one of neutral teams. They formulated their study in a
regression predicting the expected difference in score for home teams.
They set up their study to find if the home teams won games by more points when
they had home-court advantage vs.~when playing at neutral court. This study
concluded with evidence supporting home-court advantage. Also, \cite{Boudreaux} is 
able to construct a natural experiment using the Los Angeles Lakers and Los Angeles 
Clippers. Since these two teams share a home stadium, many factors like travel and 
familiarity are nullified. However, the designated home team has larger crowd support
due to attendance from their season ticket holders. When they single out the effect 
of having a sympathetic crowd, \cite{Boudreaux} estimate crowd effects
increase chance of winning 21--22.8 percentage pointts.

@@

It's worth noting that much of this paper discusses the effects from an
offensive point of view, focusing on a team's ability to score. However, one of 
the main factors that affects an offense's ability to score is the opposing
defense. It's possible that home-court advantage mainly functions as an extra
defender for home teams, which is how it negatively impacts away scoring. There
is not much home fans can do to help their team on offense, except being quiet
so players can easily communicate and focus. However, when on defense, fans can
act as a "sixth defender" supporting their team by cheering loudly and making it
harder on the offense. Also, as seen in the study \cite{Price}, fans can 
create turnovers by pressuring officials to make calls in their favor.  
This would explain how away teams were able to improve
offensively in the bubble. Home teams did not have the regular help from the
noise of fans that hinders the away team's offensive efforts. This would also
explain why home team offense didn't show significant change in the bubble, the
in-game offensive environment was largely unaltered for home teams since the
bubble provided the same quiet environment they're used to at home while on
offense. I still believe home-court advantage is better categorized as a
negative effect on away team offense, since it's more likely
noise from fans lowers level of play of away offense rather than raises the level
of play of home defense. This will be further discussed below.
